--- 
title: "GTFS Palermo"
author: "Patrick Hausmann"
output:
  html_document:
    css: style.css
    fig_caption: yes
    highlight: tango
    keep_md: yes
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 3
  github_document:
    html_preview: yes
    toc: yes
    toc_depth: 3
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE, knitr.table.format = 'markdown')
```

# Load some libraries

```{r Setup, message=FALSE, warning=FALSE}

library('broom')
library('classInt')
library('dplyr')
library('ggplot2')
library('gplots')
library('gtfsr')
library('httr')
library('leaflet')
library('maptools')
library('rbokeh')
library('readr')
library('rgdal')
library('rgeos')
library('sp')
library('tidyr')
library('viridis')

```

```{r fun}

amat_url <- "https://github.com/openamat/Official-Data/blob/master/amat_feed_gtfs.zip?raw=true"

cl_int <- function(x) { classIntervals(x, n = 5, style = "hclust", method = "complete") }

set_date <- function(x) { as.POSIXct(paste("2016/08/01", x), tz = "CET") }

arr_t_diff <- function(x) { round(as.numeric(difftime(x, lag(x), units = "mins")), 1) }

map_upl <- function(upl, df) {
  
  brks <- print(round(cl_int(df$n_stops)$brks, 0))
  
  gg_dat <- left_join(upl, df, by=c('id'='UPL')) %>%
            mutate(gr_stops = cut(n_stops,
                                  breaks = brks,
                                  right = FALSE, 
                                  include.lowest = TRUE))

  ggplot(gg_dat) +
       aes(long, lat, group=group, fill =  gr_stops) + 
       geom_polygon() +
       geom_path(color = "white") +
       coord_map() +
       #scale_fill_viridis() + 
       ggtitle("Number of stations - Le Unità di Primo Livello (UPL)")

}

map_buf <- function(dat_poly, dat_path, dat_point, title = NULL) {
  
  gg <- ggplot() + 
         geom_polygon(data = dat_poly, aes(x=long, y=lat, group=group), 
                      fill = "grey75", color = "white") +  
         geom_path(data = dat_path, aes(x=long, y=lat, group=group), 
                      color = "blue", alpha = 0.5) + 
         geom_point(data = dat_point, aes(x=longitude, y=latitude), 
                      color="red", size = 0.2) + 
         coord_map() + 
         ggtitle(title)

  return(gg)

}

```

# Import the GTFS file

```{r Import_web, eval = TRUE}

load_amat <- function(x) {
  
  if (!dir.exists("amat_gtfs")) {
    dir.create("amat_gtfs")
  }
  
  if (!file.exists(file.path("amat_gtfs", "amat.rdata"))) {
    
    message("AMAT GTFS file does not exist locally.\nStart download from GitHub 'openamat/Official-Data'")
      z <- httr::GET(x, write_disk(file.path("amat_gtfs", "amat_feed_gtfs.zip")))
      stop_for_status(z)
      amat <- import_gtfs(file.path("amat_gtfs", "amat_feed_gtfs.zip"), local = TRUE)
      save(amat, file = file.path("amat_gtfs", "amat.rdata"))
      return(amat)
    
  } else {
    message("Load local AMAT GTFS file")
      load("amat_gtfs/amat.rdata")
      return(amat)
  }
}  

amat <- load_amat(amat_url)

```

# Match UPL

```{r match_upl}

match_UPL <- function(stop_df, shp) {

    s_df <- tbl_df(stop_df) %>%
                rename(longitude = stop_lon, 
                       latitude  = stop_lat) %>%
                as.data.frame()

    coordinates(s_df) <- ~longitude+latitude
    proj4string(s_df) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
    out <- over(s_df, shp, fn = NULL)
    return(out)
}

pmo <- readOGR(dsn = "../dati/comunePalermo/UPL", 
               layer = "UPL",
               stringsAsFactors = FALSE, 
               verbose = FALSE)

proj4string(pmo)

z <- match_UPL(amat$stops_df, pmo)

amat$upl <- tbl_df(cbind(stop_id = amat$stops_df$stop_id, z))

# Number of stops
uplstops <- amat$upl %>% 
              mutate(UPL = as.numeric(UPL)) %>% 
              group_by(UPL, nome) %>% 
              summarise(n_stops = n()) %>%
              arrange(desc(n_stops))

knitr::kable(head(uplstops))

# geoDF for ggplot
pmom <- broom::tidy(pmo, region = "UPL")
pmom$id <- as.numeric(pmom$id)

```

# Number of stations per UPL

```{r map_upl,  fig.width=7, fig.height=7}

map_upl(pmom, uplstops)

```

# Import Population 2011

```{r read_upl_2011}

read_upl <- function(df){

  x <- read_csv(df) %>%
       arrange(UPL) %>%
       as.data.frame()

  # set new colnames
  y0 <- sprintf("%02.0f", seq(0, 70, by = 5))
  y1 <- sprintf("%02.0f", seq(5, 75, by = 5))
  y <- paste(y0, y1, sep = "u")
  y <- paste0("a", y)
  y <- c("UPL", "UPLDESCRIZIONE", y, "a75u", "tot")
  colnames(x) <- y

  # Mean Age
  z <- seq(2, 80, 5) + 0.5
  x$m_age <- round(rowSums(x[, 3:18] * z) / x$tot, 1)

  return(x)

}

xupl <- read_upl("../dati/comunePalermo/UPL/Popolazione per Upl al censimento 2011.csv")

knitr::kable(head(xupl))

summary(xupl$m_age)

```

# Mean age for UPL regions

```{r map_mean_age, fig.width=7, fig.height=7}

# merge data
pmom <- left_join(pmom, xupl, by=c('id'='UPL'))

ggplot() +
  geom_polygon(data = pmom, 
               aes(x=long, y=lat, group=group, fill = m_age), color = "white") +
  coord_map() +
  scale_fill_viridis() +
  ggtitle("Mean age 2011 - Le Unità di Primo Livello (UPL)")

```

# Prepare the data, direction_id = '1'

```{r}

agg_wait <- function(amat, serviceid, directionid, twindow = c(0:23)) {

  z <- tbl_df(amat$trips_df) %>%
       filter(service_id == serviceid, direction_id %in% directionid) %>%
       select(trip_id) %>%
       distinct()
  
  z1 <- amat$stop_times_df[amat$stop_times_df$trip_id %in% z$trip_id, ]
  
  z1$ix <- as.numeric(substring(z1$arrival_time, 1, 2))
  
  z1 <- tbl_df(z1) %>%
         filter(ix %in% twindow) %>%
         mutate(arrival_time = set_date(arrival_time)) %>% 
         arrange(arrival_time, stop_id)
  
  z2 <- z1 %>%
        mutate(tb = as.POSIXct(cut(arrival_time, breaks = "60 min"))) %>% 
        group_by(stop_id) %>%
        mutate(t_diff = arr_t_diff(arrival_time)) %>%
        ungroup() %>%
        arrange(stop_id, arrival_time)

  return(z2)

}

```

## Feriale August - monday-saturday

```{r}

x_FR <- agg_wait(amat, serviceid = "FR_merged_80005", directionid = 1)
x_FR <- x_FR %>% 
        group_by(stop_id) %>%
        summarise(n = n(),
                  x_mean = round(mean(t_diff, na.rm = TRUE), 1)) %>%
        ungroup()
head(x_FR) %>% knitr::kable()
sum(x_FR$n)

# http://indicators.report/indicators/i-67/
st_20min <- filter(x_FR, x_mean <= 20) 
st_20min

```

## raster (WIP)

```{r raster_plot, fig.width=7, fig.height=7}

df <- na.omit(x_FR)

stops <- tbl_df(amat$stops_df) %>%
                dplyr::select(stop_id, stop_lat, stop_lon, stop_name)

xp <- inner_join(stops, df, by = "stop_id") %>%
      rename(longitude = stop_lon, 
             latitude  = stop_lat) %>%
      as.data.frame()

coordinates(xp) <- ~longitude+latitude

rast <- raster::raster(nrows = 20, ncols = 20, 
               xmn = 13.25,
               xmx = 13.38,
               ymn = 38.06,
               ymx = 38.22,
               crs = "+proj=utm +zone=32 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")

rp <- raster::rasterize(xp, rast, field=xp$n, fun = 'sum')

plot(rp, legend.width = 1, legend.shrink = 1)
points(xp, pch = 21, cex = sqrt(xp$n/100))

```

### Cluster

```{r heatmap, eval = FALSE, include = FALSE, fig.width=9, fig.height=7}

heatm <- function(x, max_mean) {

  ix <- unique(x[x$x_mean > max_mean & 
                 !is.na(x$x_mean), "stop_id"])$stop_id
  x <- x[!x$stop_id %in% ix, ]
  x$n <- NULL
  x <- spread(x, stop_id, x_mean)
  x <- data.frame(x)
  rownames(x) <- substring(x[, 1], 12, 16)
  x <- x[, -1]
  x[is.na(x)] <- 0
  heatmap.2(as.matrix(x), 
             col = "bluered",
             Rowv = FALSE,
             dendrogram = "column",
             trace = "none") -> h1
  return(h1)

}

x_FRh <- agg_wait(amat, serviceid = "FR_merged_80005", directionid = 1)
x_FRh <- x_FRh %>% 
        select(stop_id, tb, t_diff) %>%
        group_by(stop_id, tb) %>%
        summarise(n = n(),
                  x_mean = round(mean(t_diff, na.rm = TRUE), 1)) %>%
        ungroup()
h1 <- heatm(x_FRh, max_mean = 60)

```

## Festivo August - sunday

```{r}

x_FE <- agg_wait(amat, serviceid = "FE_merged_80006", directionid = 1)
x_FE <- x_FE %>% 
        group_by(stop_id) %>%
        summarise(n = n(),
                  x_mean = round(mean(t_diff, na.rm = TRUE), 1)) %>%
        ungroup()
head(x_FE) %>% knitr::kable()
sum(x_FE$n)

```

# Number of stops per station and "waiting time" in minutes (mean)

```{r fn_freq_map}

freq_map <- function(amat, df) {

  stops <- amat$stops_df %>% 
            select(stop_id, stop_lat, stop_lon, stop_name)
  
  z <- inner_join(stops, df, by = "stop_id")
  z <- sp::SpatialPointsDataFrame(cbind("longitude" = z$stop_lon, 
                                         "latitude" = z$stop_lat), z)
  
  popup <- paste0("Stop_id: ", z$stop_id, "<br>",
                  " Stazione: ", z$stop_name, "<br>",
                  " No_of_stops: ", z$n, "<br>", 
                  " Waiting time in min (mean): ", z$x_mean)

  leaflet(z) %>% 
       setView(lng = 13.37, lat = 38.14, zoom = 12) %>%
       addTiles() %>% 
       addCircleMarkers(
         radius = ~sqrt(x_mean),
         color = "blue",
         stroke = FALSE, 
         fillOpacity = 0.5,
         popup = popup)
}

```

## Feriale - 0-24

```{r map_fr, fig.width=9, fig.height=7}

freq_map(amat, x_FR)

```

## Feriale - 18-24

```{r map_fe_7_19, fig.width=9, fig.height=7}

x_FRt <- agg_wait(amat, serviceid = "FR_merged_80005", directionid = 1, twindow = c(18:23))
x_FRt <- x_FRt %>% 
        group_by(stop_id) %>%
        summarise(n = n(),
                  x_mean = round(mean(t_diff, na.rm = TRUE), 1)) %>%
        ungroup()
freq_map(amat, x_FRt)

```

## Festivo

```{r map_fe, fig.width=9, fig.height=7}

freq_map(amat, x_FE)

```

# Cluster 'waiting time'

```{r cl, eval = TRUE, cache = FALSE}

(cl_1 <- cl_int(x_FR$x_mean))
plot(cl_1, pal = c("wheat1", "red3"))

(cl_2 <- cl_int(x_FE$x_mean))
plot(cl_2, pal = c("wheat1", "red3"))

```

# Compare

```{r com, eval = TRUE}

z <- left_join(x_FR, x_FE, by="stop_id")
head(z) %>% knitr::kable()

```

# Hexbin

```{r hexbin}

pl_hexbin <- function(amat, df, weight) {

  df <- na.omit(df)

  if (weight == "n") {
      w = round(df$n, 0)
  } else {
      w = round(df$x_mean, 0)  
  }
  
  stops <- tbl_df(amat$stops_df) %>%
             select(stop_id, stop_lat, stop_lon, stop_name)
  z <- inner_join(stops, df, by = "stop_id")
  
  lon <- rep(z$stop_lon, w)
  lat <- rep(z$stop_lat, w)
  x <- data.frame(lon, lat, 1)
  x <- sp::SpatialPointsDataFrame(cbind("longitude" = x$lon,
                                        "latitude"  = x$lat), x)
  
  figure(width = 650, height = 600) %>% 
    ly_hexbin(x$longitude, x$latitude)
}

```

## by stops

```{r hexbin_stops, fig.width=9, fig.height=7}

pl_hexbin(amat, x_FR, weight = "n")

```

## by 'waiting time'

```{r hexbin_wait, fig.width=9, fig.height=7}

pl_hexbin(amat, x_FR, weight = "x_mean")

```

# Adding buffers (WIP)

## Around each station (200m)

```{r calc_buffer, fig.width=9, fig.height=9}

add_buf <- function(dat, bufWidth) {
 
  # http://jwhollister.com/iale_open_science/2015/07/05/03-Spatial-Data-In-R/  
  d <- SpatialPointsDataFrame(coords = dat[, c("longitude", "latitude")],
                              data = dat,
                              proj4string = CRS("+init=epsg:4326"))
  
  # http://gis.stackexchange.com/a/136285
  # reproject to Google Mercator (meters)
  d_mrc <- spTransform(d, CRS = CRS("+init=epsg:3857"))

  # see for 'byid' 
  # http://bit.ly/2aGi0MP
  d_bff <- gBuffer(d_mrc, byid = TRUE, width = bufWidth)
  d_bff <- SpatialPolygonsDataFrame(d_bff, d@data)
  d_bff@data$BufferWidthM <- bufWidth
  
  # reproject to ...
  d_bff_rp <- spTransform(d_bff, CRS("+init=epsg:4326"))
  d_bff_rp_ft <- broom::tidy(d_bff_rp)
  
  return(d_bff_rp_ft)
  
}

dat <- tbl_df(amat$stops_df) %>%
         select(stop_lon, stop_lat, stop_id, stop_name) %>%
         rename(longitude = stop_lon, latitude = stop_lat) %>%
         as.data.frame()

buf_dat <- add_buf(dat, 200)

```

```{r buffer_all, eval = TRUE, fig.width=9, fig.height=9}

map_buf(dat_poly = pmom, 
        dat_path = buf_dat,
        dat_point = dat, 
        title = "200m Buffer around each station")

```

## Around stations with a transit frequency <= 20min

NOT sure if correct!

```{r buffer_sp20, eval = TRUE, fig.width=9, fig.height=9}

# http://indicators.report/indicators/i-67/
buf_dat_st20 <- filter(buf_dat, id %in% st_20min$stop_id)

map_buf(dat_path = buf_dat_st20,
        dat_poly = pmom, 
        dat_point = dat, 
        title = "200m Buffer around each station, transit frequency <= 20min")

```

```{r rsession}

devtools::session_info()

```
